---
title: 'rSMILE for pharma risk model '
authors: Miriam Glendell, Camilla Negri & Mads Troldborg
date: "2025-07-01"
output:
  pdf_document: default
  html_document: default
---

Notebook for running the pharma risk model spatially using rSMILE.

**Important note:**  
This analysis requires the proprietary **GeNIe** software and the R package
`rSMILE`. Due to licensing restrictions, the BN model file is not included in
this repository. This document is provided for transparency and reproducibility
of the modelling workflow.

## Data inputs
This analysis uses a combination of:

**(a) Shareable, processed data**
- Monthly prescribing statistics (values derived in 01_process_prescribing_data.R and built in to the BN model)
- Flow statistics by monitoring point (values derived in 03_process_flow_data.R and saved in Data/processed)
- Sewage treatment work (STW) information summarised by monitoring point sub-catchments (derived in 02_process_WWTW_info.R and saved in Data/processed) 
- Excretion parameters (derived in 04_process_excretion_data.Rmd and built into BN model)
- Wastewater removal (derived in 05_process_STW_removal_efficiencies.Rmd and built into BN model)  
- Bayesian Network model file (`.xdsl`)  

**(b) Restricted / licensed inputs**
- Observed concentration datasets
- rSMILE / GeNIe license

Restricted inputs are expected to be located in `Data_private/`.


## Outputs:
 - Data/processed/simulated_PEC_by_pharma_xxxTreatment_xxxPrescribing.RData
 - Outputs/figures/plot_ng_XXX_xxxTreatment_xxxPrescribing.png
 - Outputs/figures/plot_EQR_MP_XXX_xxxTreatment_xxxPrescribing.png
 - Outputs/summary_all_pharma_PEC_xxxTreatment_xxxPrescribing.csv
 - Outputs/results_EQR_all_MPs_xxxTreatment_xxxPrescribing.csv


# Load libraries and paths
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!requireNamespace("rSMILE", quietly = TRUE)) {
  stop(
    "rSMILE not installed. This analysis requires licensed GeNIe software ",
    "and cannot be executed without it."
  )
}

library(rSMILE)
library(stringi)
library(tidyverse)
library(scales)
library(here)

source(here::here("R", "utils.R"))

```

# Load license 
```{r}
# License issued by BayesFusion Licensing Server and must be obtained in order to run this notebook

# This code must be executed before any other rSMILE object is created 
source(file.path(path_private, "rSMILE_License.R"))

```


# Read BN model 
```{r, echo=FALSE}
net <- Network() 
net$readFile(file.path(path_model, "Pharma Model Almond.xdsl"))

```


# Load flow statisitcs 
```{r, echo=FALSE, include=FALSE}
# Mean and SD of the simulated daily discharge (using SWAT) on log-scale at each monitoring point (sub-catchment) by month. Basically, the flows at each monitoring point are (assumed to be) log-normal distributed.
Evidence_flow <- read.csv(file.path(path_processed, "monthly_mean_and_sd_simulated_daily_logQ_MP_Almond.csv"))  
Evidence_flow <- subset(Evidence_flow, select= -c(MP_name,mean_Q, median_Q, sd_Q, median_lnQ)) %>% 
  rename(MP = MP_no,
         lnQ_mean = mean_lnQ,
         lnQ_sd = sd_lnQ)

#split into a list
Discharge_scenarios <- split(Evidence_flow , f = Evidence_flow$MP)
```

# Load STW information/evidence 
```{r, echo=FALSE, include=FALSE}
# read in evidence for STW PE and treatment type for each monitoring point
treatment_PE <- read.csv(file.path(path_processed,"STW_PE_Dist_Treatment_by_MP.csv")) # 

treatment <- treatment_PE %>% dplyr::select(MP_no,SepticTank,Preliminary,Primary,SecondaryA,SecondaryB,TertiaryA1,TertiaryA2,TertiaryB1,TertiaryB2)

# transpose
treatment2 <- treatment %>% 
  rownames_to_column() %>% 
  pivot_longer(!rowname, names_to = "col1", values_to = "col2") %>% 
  pivot_wider(names_from = "rowname", values_from = "col2")

# rename first column name to 'Treatment_type'
treatment2 <- treatment2 %>% rename(Treatment_type=col1)

# remove the first column from treatment2
treatment3 <- treatment2 %>% dplyr::select(-Treatment_type)
treatment3 <- treatment3[-c(1),] # gets rid of first row 
names(treatment3) <- as.character(treatment$MP_no)
treatment3 <- round(treatment3,digits=3)


```


# Extract Total PE by monitoring point
```{r}
Total_PE_df <- dplyr::select(treatment_PE,Total_PE)
Total_PE_df <- round(Total_PE_df,0)
rownames(Total_PE_df) <- as.character(treatment$MP_no)

```


# Write flow equations for Genie
```{r}
lognormal_scenario_equations<- function(mydf){
    mean <- mydf$lnQ_mean
    sd   <- mydf$lnQ_sd
    lognormals<-
      sprintf("Lognormal(%.2f, %.3f)", mean, sd) %>%
      stri_paste(collapse=',') # string together all the lognormals
    equation <- paste0("Disch=Choose(Month,",lognormals,")")
    equation #include incipit of equation "Disch=Choose(Month," and then closing bracket
  return(equation)
}
```


# Get flow equation for every scenario (every df in the list)
```{r}
all_Discharge_equations <- lapply(Discharge_scenarios, lognormal_scenario_equations)


```

# Get monthly matrix to get monthly results 
```{r}
# apply discrete evidence on a single discrete node (month)
# set evidence on Month
months      <- str_to_lower(month.abb)
months_prob <- data.frame(diag(rep(1,12)))
colnames(months_prob) <- months

```


# Write PE equations
```{r}
# Initialize an empty list for PE equations
PE_equations <- vector("list", length = nrow(Total_PE_df))  

# Iterate over the rows of Total_PE_df
for (i in 1:nrow(Total_PE_df)) {
  # Check if the value is zero
  if (Total_PE_df[i, ] == 0) {
    # Print the row name where the value is zero
    print(paste("Zero found in item:", rownames(Total_PE_df)[i]))
  }
  
  # Create the equation for each row
  PE_equations[[i]] <- sprintf("Total_PE=%.1f", Total_PE_df[i, ])
}

# Assign row names from Total_PE_df to the list
names(PE_equations) <- rownames(Total_PE_df)
```

# Update BN model with location specific data and extract resulting simulated pharma concentration (or whatever target node is chosen).  
```{r}

calculate_posterior <- function(my_net, node1_handle, node2_handle, node3_handle, node4_handle, target_node, equations_list, probabilities_df, treatment_df, Total_PE_equations, Pharma_id, treat_mgmt="baseline", prescr_mgmt="baseline") {
  
  results <- list()  # Initialize an empty list to store the results
  
  # Set evidence for pharmaceutical compound
  my_net$setEvidence("Pharma", Pharma_id)
  
  # Set evidence for treatment management scenario
  my_net$setEvidence("Treatment_scenario", treat_mgmt) 
  
  # Set evidence for prescribing management scenario
  my_net$setEvidence("Prescription_scenario", prescr_mgmt)
 
  # Iterate over the locations (20 locations from equations_list)
  for (eq_name in names(equations_list)) {   # eq_name is the unique location number (as a string)
    
    # Get the equation for node4 from the Total_PE_equations list for this location
    node4_eq <- Total_PE_equations[[eq_name]]  # This should match the location name
    
    # If PE = 0, then there is no need to calculate the posterior
    if(node4_eq == "Total_PE=0.0") {
      month_states     <- str_to_lower(month.abb)
      location_results <- switch(target_node,
                                 PEC  = setNames(rep(list(0), 12), month_states),
                                 PLRR = setNames(rep(list(-4), 12), month_states),
                                 PRRQ = setNames(rep(list(c(1, 0)), 12), month_states),
                                 stop("Unknown Target Node")
                                 )
    } else {
      
    # Initialize a list to store results for this location (eq_name)
    location_results <- list()
    
    my_net$setNodeEquation(node4_handle, node4_eq)
    
    # Get the equation for the current location for node2
    node2_eq <- equations_list[[eq_name]]
    my_net$setNodeEquation(node2_handle, node2_eq)
    
    # Extract treatment for this location (assumed to be in the correct order)
    # The `eq_name` corresponds to a specific location, which matches a column name in the `treatment_df` tibble. The treatment values for the location is extracted using
    # `treatment_df[[eq_name]]`. This yields the treatment values for the specific location as a one-column tibble, which is then converted to a numeric vector using
    # `as.numeric()`. The treatment values are expected to be in the correct order /  number (e.g., as many values as there are treatment options).
    treatment <- as.numeric(treatment_df[[eq_name]])  # Convert to numeric vector
    
    # Set treatment for this location
    # The `my_net$setNodeDefinition(node3_handle, treatment)` sets the treatment values for the current location (specified by `eq_name`)
    my_net$setNodeDefinition(node3_handle, treatment)  # Set the treatment for this location

    # Iterate through each month (columns in probabilities_df)
    for (i in 1:ncol(probabilities_df)) {
      
      # Define node 1 using the values from the current column (month) in probabilities_df
      my_net$setNodeDefinition(node1_handle, probabilities_df[, i])  # Set node1 for this month

      # Update beliefs in the network
      my_net$updateBeliefs()
      
      # Get posterior beliefs for the target node
      beliefs <- my_net$getNodeValue(target_node)
      
      # Store beliefs for this location and month (use month names from probabilities_df columns)
      month_name <- colnames(probabilities_df)[i]
      location_results[[month_name]] <- beliefs
    }
    
    } # end of if else
    
    # Store the location results (eq_name) in the results list
    results[[eq_name]] <- location_results
  }
  
  return(results)  # Return the final results
}


```


# Model all pharmaceuticals at all locations
```{r}
# List of pharmaceuticals
pharmas<-c("AZM","BZF","CBZ","CIP","CLR","DFC","ERY","HCT","IBU","NAP","PAR","RAN","SDZ","SMX","TML","TMP")

```

# ------------------------------------- # 
# RUN MODEL WITH PEC AS TARGET NODE
# ------------------------------------- # 
```{r}
start_time <- Sys.time()

# Select only points with PE>0
id0       <- which(Total_PE_df$Total_PE>0)
id0_names <- rownames(Total_PE_df)[id0]
selected_Discharge_equations <- all_Discharge_equations[id0_names]
selected_treatment    <- treatment3[,names(treatment3) %in% id0_names]
selected_PE_equations <- PE_equations[id0]

# Initialize a list to store results for each pharma
posterior_results_by_pharma <- list()

# clear all previous evidence in model
net$clearAllEvidence()

# CHOOSE TARGET NODE 
# Target node options: PEC, PLRR, PRRQ
my_target  = "PEC" # 

# CHOOSE treatment scenario (baseline or optimal)
treat_scenario  = "baseline" # "baseline" # "optimal"
# CHOOSE prescribing scenario (baseline, reduced_10pct or reduced_50pct)
prescr_scenario = "baseline" # "reduced_50pct" # "baseline" # "reduced_10pct"

# Iterate over each pharma
for (pharma in pharmas) {
  
  # Apply calculate_posterior for each pharma
  posterior_results_by_pharma[[pharma]] <- calculate_posterior(
    my_net = net,
    node1_handle = "Month",
    node2_handle = "Disch",
    node3_handle = "Treatment_type",
    node4_handle = "Total_PE", 
    target_node  = my_target, # PEC, PLRR PRRQ # would be PLRR for EQR and PRRQ for discrete node of probability of exceedance, PEC otherwise
    equations_list = selected_Discharge_equations,#all_Discharge_equations,
    probabilities_df = months_prob,
    treatment_df = selected_treatment, 
    Total_PE_equations = selected_PE_equations, 
    Pharma_id   = pharma,  # Use pharma for Pharma_id
    treat_mgmt  = treat_scenario,
    prescr_mgmt = prescr_scenario
    )
  
}

end_time <- Sys.time()
end_time - start_time


```
<!-- # Save output -->
<!-- ```{r} -->
<!-- # save -->
<!-- saveRDS( -->
<!--   posterior_results_by_pharma, -->
<!--   file = file.path(path_processed, paste0("posterior_",my_target,"_by_pharma_list_",treat_scenario,"Treatment_",prescr_scenario,"Prescribing.rds")) -->
<!-- ) -->

<!-- ``` -->


# Create dataframes for figures 
```{r}

combined_df <- imap_dfr(posterior_results_by_pharma, ~ {
  pharma_name <- .y
  imap_dfr(.x, ~ {
    location_name <- .y
    imap_dfr(.x, ~ {
      tibble(
        Pharma        = factor(pharma_name),
        Location      = factor(paste0("MP",location_name)),
        Month         = .y,
        Concentration = .x
      )
    })
  })
})

combined_df <- combined_df %>% mutate(Month = factor(Month, levels = str_to_lower(month.abb)))

# save combined_df for later comparison to observed concentrations
save(
  combined_df,
  file = file.path(path_processed, paste0("simulated_",my_target,"_by_pharma_",treat_scenario,"Treatment_",prescr_scenario,"Prescribing.RData"))
)

rm(posterior_results_by_pharma)
```


# PLOT RESULTS
```{r}
# Define a custom palette with 20 colors because there are 20 locations, even though some will have 0 concentration
custom_colors <- c(
  "#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#e6ab02", "#a6761d", "#666666", 
  "#1f78b4", "#33a02c", "#fb9a99", "#cab2d6", "#ff7f00", 
  "#6a3d9a", "#b15928", "#a6cee3", "#b2df8a", "#fdbf6f", "#ff99ff", "#8dd3c7",
  "yellow","red","grey"
)
```


```{r}
# Increase the size of the plotting window in RStudio
options(repr.plot.width = 14, repr.plot.height = 10)

# Only plot points for which we have measurements:
selected_locations <- paste0("MP",id0_names[as.numeric(id0_names) <= 21]) # 

unique_pharmas <- unique(combined_df$Pharma)

# Iterate over each pharma
for (pharma in unique_pharmas) {
  # Extract the transformed data frame for the current pharmaceutical
 # pharma_df <- pharma_data_list[[pharma]]
  
  # Filter out rows with Concentrations that are 0
  pharma_df <- combined_df %>% filter(Pharma == pharma,
                                    Concentration != 0,
                                    Location == selected_locations)
  
  # Multiply Concentration by 1000 to scale it up to ng L-1
  pharma_df <- pharma_df %>% mutate(Concentration = Concentration * 1000)
  
  # Generate the plot using ggplot
  log10_plots_ng <- ggplot(pharma_df, aes(x = Month, y = log10(Concentration), color = Location)) + 
    geom_boxplot(width = 0.25, outlier.shape = NA) +
    scale_color_manual(values = custom_colors) +  # Apply custom palette
    labs(
      title = paste("Concentrations of", pharma),  # Customized title
      x = "Month", y = "log10 Concentration (ng/L)"  # Update y-axis label to ng/L
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    facet_wrap(~ Location, , scales = "free_y") +
    scale_y_continuous(limits = c(-4,5))  # Adjusted the y-axis limits for ng/L scale
  
  # Print the plot to display in the R session
 print(log10_plots_ng)

ggsave(
  filename=paste0("plot_ng_",pharma,"_",treat_scenario,"Treatment_",prescr_scenario,"Prescribing.png"),
  log10_plots_ng,
  path = file.path(path_figures),
  width = 10,#180,
  height = 6 #140,
  #units = "mm",
  #dpi = 600
)

}


```


# Get statistics from results of continuous variables (results in ug L-1 on real scale, as in the model)
```{r}

# Group by Pharma, Location, and Month, and calculate summary statistics
summary_df <- combined_df %>%
  group_by(Pharma, Location, Month) %>%
  summarise(
    mean_conc = mean(Concentration, na.rm = TRUE),  # Mean of concentrations 
    sd_conc = sd(Concentration, na.rm = TRUE),  # Standard deviation 
    n = n(),  # Number of data points
    median_conc = median(Concentration, na.rm = TRUE)  # Median of concentrations 
  ) %>%
  mutate(
    # Calculate the lower and upper limits based on the log mean and SD
    low_conc = mean_conc - sd_conc,
    upper_conc = mean_conc + sd_conc,
    ) %>%
  dplyr::select(Pharma, Location, Month, mean_conc, sd_conc, median_conc, n, low_conc, upper_conc)


```
# Add statistics for location with PE=0
```{r}

# Get locations with PE==0
id0_location <- paste0("MP", rownames(Total_PE_df)[which(Total_PE_df$Total_PE==0)])
summary_df_add <- expand.grid(Pharma=pharmas, Location=id0_location, Month=str_to_lower(month.abb), mean_conc=0, sd_conc=0, median_conc=0,n=0,low_conc=0,upper_conc=0)

summary_df_add <- summary_df_add[with(summary_df_add, order(Pharma,Location)),]
summary_df     <- rbind(summary_df,summary_df_add)

```

# save statistics
```{r}

write.csv(
  summary_df,
  file = file.path(path_outputs, paste0("summary_all_pharma_", my_target,"_",treat_scenario,"Treatment_",prescr_scenario,"Prescribing.csv")),
  row.names = FALSE
)
  

```


# ------------------------------------- # 
# RUN MODEL WITH PRRQ AS TARGET NODE
# ------------------------------------- # 
```{r}
start_time <- Sys.time()

# Select all points with PE>=0
id0       <- which(Total_PE_df$Total_PE>=0)
id0_names <- rownames(Total_PE_df)[id0]
selected_Discharge_equations <- all_Discharge_equations[id0_names]
selected_treatment    <- treatment3[,names(treatment3) %in% id0_names]
selected_PE_equations <- PE_equations[id0]

# Initialize a list to store results for each pharma
posterior_results_by_pharma <- list()

# clear all previous evidence in model
net$clearAllEvidence()

# CHOOSE TARGET NODE 
# Target node options: PEC, PLRR, PRRQ
my_target  = "PRRQ" # 

# CHOOSE treatment scenario (baseline or optimal)
treat_scenario  = "baseline" # "baseline" # "optimal"
# CHOOSE prescribing scenario (baseline, reduced_10pct or reduced_50pct)
prescr_scenario = "baseline" # "reduced_50pct" # "baseline" # "reduced_10pct"


# Iterate over each pharma
for (pharma in pharmas) {
  
  # Apply calculate_posterior for each pharma
  posterior_results_by_pharma[[pharma]] <- calculate_posterior(
    my_net = net,
    node1_handle = "Month",
    node2_handle = "Disch",
    node3_handle = "Treatment_type",
    node4_handle = "Total_PE", 
    target_node  = my_target, # PEC, PLRR PRRQ # would be PLRR for EQR and PRRQ for discrete node of probability of exceedance, PEC otherwise
    equations_list = selected_Discharge_equations,#all_Discharge_equations,
    probabilities_df = months_prob,
    treatment_df = selected_treatment, 
    Total_PE_equations = selected_PE_equations, 
    Pharma_id   = pharma,  # Use pharma for Pharma_id
    treat_mgmt  = treat_scenario,
    prescr_mgmt = prescr_scenario
    )
  
}

end_time <- Sys.time()
end_time - start_time

```


# Save output
```{r}
# save
saveRDS(
  posterior_results_by_pharma, 
  file = file.path(path_processed, paste0("posterior_",my_target,"_by_pharma_list_",treat_scenario,"Treatment_",prescr_scenario,"Prescribing.rds"))
) 






```


# Dataframes for bar charts to plot EQR figures
```{r}
# Initialize a list to store the data frames for each pharmaceutical
pharma_data_list2 <- list()

# Loop over each pharmaceutical in posterior_results_by_pharma
for (pharma in names(posterior_results_by_pharma)) {
  # Initialize an empty list to store each location's data for the current pharma
  pharma_location_data <- list()
  
  # Loop over each location within the current pharma
  for (location in names(posterior_results_by_pharma[[pharma]])) {
    # Extract the data for each month at the current location and pharma
    location_data <- posterior_results_by_pharma[[pharma]][[location]]
        
    if(location=="21") location="20"  # CHANGING NAME OF MP21 TO MP20
    
    # Convert location_data to a data frame in long format for ggplot, adding Pharma and Location
    location_df <- data.frame(
      Month = factor(rep(names(location_data),each=2)),  # Month as factor
      EQR   = factor(seq(1,2)),
      Probability = unlist(location_data),
      Location = factor(paste0("MP", location)),  # Location as "MP1", "MP2", etc., and as factor
      Pharma = factor(pharma)  # Pharma as factor
   ) 
    levels(location_df$EQR)<-c("Below","Above")
   
    # Append the location data to the current pharma's list
    pharma_location_data[[location]] <- location_df
  }
  
  # Combine all location data frames for the current pharma
  pharma_data_list2[[pharma]] <- do.call(rbind, pharma_location_data)
  rownames(pharma_data_list2[[pharma]])<-seq(1:nrow(pharma_data_list2[[pharma]])) #fix a problem with row naming
}
```


# Plotting discrete target variable PRRQ

```{r}
# plotting results of discrete target variable PRRQ
#Increase the size of the plotting window in RStudio
options(repr.plot.width = 14, repr.plot.height = 10)

select_MPs <- unique(pharma_data_list2[[1]]$Location)

# Use `pharma_data_list2` for plotting discrete probabilities
for (pharma in names(pharma_data_list2)) {
  # Extract the transformed data frame for the current pharmaceutical
  pharma_df <- pharma_data_list2[[pharma]] %>% filter(Location %in% select_MPs)
  
  # Generate the plot using ggplot
  EQR_plots <- ggplot(pharma_df, aes(x = fct_inorder(Month), y=Probability, fill = EQR)) + 
    geom_col(position="stack", colour = "black",linewidth = 0.15, width = 1) +  
    scale_fill_manual(values = c("Below" = "grey80",  "Above" = "grey30")) +
    #scale_fill_manual(values = c("Below" = "#1E88E5", "Above" = "#FFC107")) +  # colour option
    scale_y_continuous(
       labels = percent_format(accuracy = 1),
       expand = c(0, 0)
       ) +    
    facet_wrap(~ Location, ncol=5) +
    labs(
      title = paste("Probability of exceedance", pharma),  # Customized title
      x = "Month", y = "Probability", fill = "EQR" 
    ) + 
  
  theme_bw(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 90, vjust=0.5, hjust = 1),
    legend.position = "top"
  )


  # Print the plot to display in the R session
  print(EQR_plots)
  
  ggsave(
  filename=paste0("plot_EQR_MP_",pharma,"_",treat_scenario,"Treatment_",prescr_scenario,"Prescribing.png"),
  EQR_plots,
  path = file.path(path_figures),
  width = 220,
  height = 140,
  units = "mm",
  dpi = 600
)

}
```


```{r}
# Now combine all pharmaceuticals into a single data frame and save in csv file
combined_df_2 <- do.call(rbind, pharma_data_list2)

write.csv(
  combined_df_2,
  file = file.path(path_outputs, paste0("results_EQR_all_MPs_",treat_scenario,"Treatment_",prescr_scenario,"Prescribing.csv")),
  row.names = FALSE
)
  
```



